---
title: "data clean"
author: "Huiyi Zhu"
date: "2025-12-04"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## prepare

```{r}
library(tidyverse)
library(janitor)
library(ggplot2)
library(patchwork)
```
outcome: attendance 
variables: season, event type, borough, location type, time of date



Miranda Chen: model (before 11:59PM Friday)
Huiyi Zhu: clean data (date:2022.5-present, morning5-12 afternoon 12-6 night 6-1 midnight?) + basic descriptive analysis 
Yiyang Jiao: report+video (9:00PM Sat)
Xiaoqian Lyu: report
Yuyue Ma: web(sat) 


## Data Cleaning
```{r}
parkevent = read.csv("Parks_Special_Events_20251204.csv") |>
  janitor::clean_names() |>
  mutate(
    datetime = mdy_hms(date_and_time),
    date     = as_date(datetime), 
    year  = year(datetime),
    month = month(datetime, label = TRUE, abbr = TRUE),
    day = day(datetime),
    time = format(datetime, "%H:%M:%S"),
    season = case_when(
      month %in% c("Mar","Apr","May") ~ "Spring",
      month %in% c("Jun","Jul","Aug") ~ "Summer",
      month %in% c("Sep","Oct","Nov") ~ "Fall",
      month %in% c("Dec","Jan","Feb") ~ "Winter"), 
      season = factor(season,
        levels = c("Spring", "Summer", "Fall", "Winter")),
    hour = as.numeric(format(datetime, "%H")),
    time_period = case_when(
      hour < 5 ~ "Midnight",
      hour >= 5  & hour < 12 ~ "Morning",
      hour >= 12 & hour < 18 ~ "Afternoon",
      hour >= 18 & hour < 23 ~ "Night",
      hour >= 23   ~ "Midnight"),
   time_period = factor(time_period,
                     levels = c("Morning", "Afternoon", "Night", "Midnight")),
    attendance = as.numeric(attendance))|>  
  select(
    -datetime,
    -unit,
    -group_name_partner,
    -date_and_time,
    -classification,
    -source,
    -audience,
    -hour,
    -event_name,
    -location) |> 
  select(
    date, year, month, day, time, season,
    everything()) |> 
    filter(date >= as.Date("2021-06-01"),
           event_type != "",
           category != "",
           attendance != "")
```

We cleaned the dataset to prepare it for analyzing attendance as the primary outcome and key predictors including season, event_type, borough, location_type, and time_period. Date-time information was parsed into date, year, month, and day to support temporal analyses, and time_period was created to evaluate differences across times of day. We filtered observations to the NYC fully reopened period after COVID (date >= "2021-06-01") to avoid pandemic-related disruptions. Columns not relevant to the analysis were removed, and rows with empty event_type, category and attendance values were dropped. After cleaning, we have `r nrow(parkevent)` observations and `r ncol(parkevent)` variables. These steps produced a tidy dataset aligned with our study.



## Descriptive Analysis 
```{r}
parkevent |>
  summarize(
    mean_attendance = mean(attendance),
    median_attendance = median(attendance),
    sd_attendance = sd(attendance),
    min_attendance = min(attendance),
    max_attendance = max(attendance)) |>
  knitr::kable(digits = 2)
```

This table summarizes the distribution of attendance across all post-COVID events, including the mean, median, variability, and range of turnout. These descriptive statistics provide an essential baseline for understanding overall engagement levels and for comparing attendance across seasons, event types, boroughs, and time-of-day categories in subsequent analyses.

```{r}
plot_season =
  ggplot(parkevent, aes(x = season, y = attendance, fill=season)) +
  geom_violin(trim = FALSE, alpha = 0.5) +
  geom_boxplot(width=0.15) +
  coord_cartesian(ylim = c(0,250)) +
  scale_x_discrete(labels = c(
    "Spring" = "Spr",
    "Summer" = "Sum",
    "Fall"   = "Fal",
    "Winter" = "Win")) +
  labs(title = "By Season", x = "Season", y = "Attendance") +
  guides(fill = "none") +
  theme_minimal()

plot_time =
  ggplot(parkevent, aes(x = time_period, y = attendance, fill=time_period)) +
  geom_violin(trim = FALSE, alpha = 0.5) +
  geom_boxplot(width=0.15) +
  coord_cartesian(ylim = c(0,250)) +
  scale_x_discrete(labels = c(
    "Morning"   = "Morn",
    "Afternoon" = "Aft",
    "Night"     = "Ngt",
    "Midnight"  = "Mid")) +
  labs(title = "By Time of Day", x = "Time of Day", y = "Attendance") +
  guides(fill = "none") +
  theme_minimal()

plot_type =
  ggplot(parkevent, aes(x = event_type, y = attendance, fill = event_type)) +
  geom_violin(trim = FALSE, alpha = 0.5) +
  geom_boxplot(width = 0.15) +
  coord_cartesian(ylim = c(0,250)) +
  scale_x_discrete(labels = c(
    "Agency Produced Event" = "Agency",
    "Community Based Event" = "Community",
    "Local Event" = "Local",
    "Open House" = "Open",
    "Tournament/Competition" = "Compet")) +
   guides(fill = "none") +
  labs(
    title = "By Event Type",
    x = "Event Type",
    y = "Attendance") +
  theme_minimal()

plot_borough =
  ggplot(parkevent, aes(x = borough, y = attendance, fill=borough)) +
  geom_violin(trim = FALSE, alpha = 0.5) +
  geom_boxplot(width=0.15) +
  coord_cartesian(ylim = c(0,250)) +
  scale_x_discrete(labels = c(
    "Bronx"        = "BX",
    "Brooklyn"     = "BKN",
    "Manhattan"    = "MAN",
    "Queens"       = "QNS",
    "Staten Island"= "SI")) +
  labs(title = "By Borough", x = "Borough", y = "Attendance") +
  guides(fill = "none") +
  theme_minimal()

plot_location =
  ggplot(parkevent, aes(x = location_type, y = attendance, fill = location_type)) +
  geom_violin(trim = FALSE, alpha = 0.5) +
  geom_boxplot(width = 0.15) +
  coord_cartesian(ylim = c(0,250)) +
  scale_x_discrete(labels = c(
    "Nature Center" = "Nature",
    "Recreation Center" = "RecCen",
    "Playground" = "Playground",
    "Pool" = "Pool",
    "Park" = "Park",
    "Other" = "Other",
    "School" = "School")) +
  labs(
    title = "By Location Type",
    x = "Location Type",
    y = "Attendance") +
   guides(fill = "none") +
  theme_minimal()

patterns_plot =
  ((plot_season |plot_time| plot_borough) /
    (plot_type| plot_location))+
  plot_annotation(
    title = "Attendance Patterns Across NYC Parks Events",
    theme = theme())
patterns_plot

```

```{r}
plot_time_season =
  parkevent |>
  group_by(time_period, season) |>
  summarize(mean_attendance = mean(attendance)) |>
  ggplot(aes(x = time_period, 
             y = mean_attendance, 
             color = season, 
             group = season)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Mean Attendance by Time of Day Across Seasons",
    x = "Time of Day",
    y = "Mean Attendance",
    color = "Season") +
  theme_minimal()

plot_borough_season =
  parkevent |>
  group_by(borough, season) |>
  summarize(mean_attendance = mean(attendance)) |>
  ggplot(aes(x = borough, 
             y = mean_attendance, 
             color = season, 
             group = season)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Mean Attendance by Borough Across Seasons",
    x = "Borough",
    y = "Mean Attendance",
    color = "Season") +
  theme_minimal()

Seasonal_plot =
  (plot_time_season / plot_borough_season) +
  plot_annotation(
    title = "Seasonal Attendance Patterns Across NYC Park Events")

Seasonal_plot
```



```{r}

plot_calendar =
  ggplot(parkevent, 
         aes(x = month(date, label = TRUE, abbr = TRUE), 
             y = attendance)) +
  geom_point(alpha = 0.2, color = "gray") +
  stat_summary(fun = mean, geom = "line", aes(group = 1),
               color = "black", size = 1) +
  stat_summary(fun = mean, geom = "point",
               color = "black", size = 2) +
  labs(
    title = "Attendance Over the Calendar Year",
    x = "Month",
    y = "Attendance") +
  coord_cartesian(ylim = c(0, 250)) + 
  theme_minimal()

plot_calendar
```



